Quantifying stochastic noise in cultured\\ circadian reporter cells

### 2015-03-12 main_text.tex
another edit from frank
further updates, adding FigS6

\pagebreak
\begin{abstract}
Stochastic noise at the cellular level has been shown to play a fundamental role in circadian oscillations, influencing how groups of cells entrain to external cues and likely serving as the mechanism by which cell-autonomous rhythms are generated.
Despite this importance, few studies have investigated how clock perturbations affect stochastic noise - even as increasing numbers of high-throughput screens categorize how gene knockdowns or small molecules can change clock period and amplitude.
This absence is likely due to the difficulty associated with measuring cell-autonomous stochastic noise directly, which requires the careful collection and processing of single-cell data.
% since cultured circadian reporters lack significant intercellular coupling,
In this study, we show that the damping rate of population-level bioluminescence recordings can serve as an accurate measure of cell-autonomous stochastic noise, and one which can be applied to future and existing high-throughput circadian screens.
Using cell-autonomous fibroblast data, we first show directly that higher noise at the single-cell results in faster damping at the population level.
Next, we show that the damping rate of cultured cells can be changed in a dose-dependent fashion by small molecule modulators, and confirm that such a change can be explained by single-cell noise using a mathematical model.
We further demonstrate the insights that can be gained by applying such a method to a genome-wide siRNA screen, revealing that stochastic noise is altered independently from period, amplitude, and phase.
Finally, we hypothesize that the unperturbed clock may be highly optimized for robust rhythms, as very few gene perturbations are capable of simultaneously increasing amplitude and lowering stochastic noise.
Ultimately, this study demonstrates the importance of considering the effect of circadian perturbations on cell-autonomous noise, particularly with regard to the development of small-molecule circadian therapeutics.
\end{abstract}

@@

Circadian rhythms are daily changes in gene expression and physiology that persist even in the absence of external environmental cues \cite{Herzog2007}.
In mammals, such rhythms are organized in a hierarchical fashion.
At the tissue-level, the brain's suprachiasmatic nucleus (SCN) serves as the master pacemaker and keeps circadian oscillations in peripheral tissues in phase with the light-dark cycle.
Cell-to-cell coupling in the SCN keeps individual cells in tight synchrony \cite{Herzog2004}. 
Within each tissue, cellular-level rhythms are generated by a large network of interacting regulatory elements, in which time-delayed negative feedback gives rise to sustained oscillations in gene transcription \cite{Ueda2005}.
The robust oscillation of circadian factors has been linked to metabolic health \cite{Bass2012}, as compromised rhythms, whether due to gene knockout \cite{Marcheva2010} or irregular feeding schedules \cite{Hatori2012}, result in an increased risk of metabolic disease.
Additionally, since the amplitude of circadian transcription can be affected by lifestyle variables such as diet, age, or work schedules, there has been recent interest in developing pharmacological strategies for increasing the amplitude of circadian cycles in metabolic tissues \cite{Chen2013}.

A detailed understanding of the underlying transcriptional mechanisms is essential for the development of circadian therapeutics to be successful. 
The functional roles of different species in circadian regulation have traditionally been studied using behavioral-level data and genetic knockout experiments \cite{Vitaterna1994}.
Bioluminescence-based cellular circadian reporters offer a more direct view of the gene regulatory network \cite{Balsalobre1998} and are amenable to high-throughput screens, allowing genome-wide exploration into factors which affect circadian rhythmicity \cite{Zhang2009}.
Additionally, cultured circadian reporter cells allow the change in transcriptional amplitude following a perturbation to be quantified.
This additional parameter has proven useful in differentiating between perturbations with the same effect on period \cite{St.John2014}, and has aided the search for small-molecule therapeutics to boost clock amplitude \cite{Chen2013}.

@@

Despite the averaging which occurs at the population-level, cell-autonomous stochastic noise plays an important role in determining the function of the circadian oscillator.
A recent study showed that stochasticity is critical to the population-level response to a neuropeptide and forms the basis for how the SCN entrains to light-mediated cues \cite{An2013}.
Additional studies have argued that the basis of single-cell rhythmicity depends on stochastic noise, as models of deterministically damped oscillators, when simulated stochastically, better capture the noise characteristics seen in single-cell fibroblast data \cite{Westermark2009}.
Despite the importance of single-cell stochasticity in circadian rhythms, measuring stochastic noise currently requires careful preparation and recording of individual cells and subsequent image processing \cite{Leise2012}. 
As a result, while circadian perturbations have been postulated to affect single-cell stochasticity \cite{Rougemont2007}, no study has experimentally quantified changes to single-cell noise as a result of a small molecule or genetic perturbation.

In this study, we demonstrate that changes to single-cell stochasticity can be reliably inferred from the changes to the damping rate of population-level bioluminescence recordings of cultured circadian reporters.
Our method assumes the following conditions are met, which have been shown to hold experimentally for many types of cultured cells \cite{Nagoshi2004, Welsh2004, Guenthner2014, Noguchi2013}:
oscillations in individual cells are
\begin{itemize}
  \item independent (no significant cell-to-cell coupling),
  \item sustained (do not damp on a single-cell basis),
  \item and homogeneous, with no differences in expected period length between cells.
\end{itemize}
We demonstrate the validity and usefulness of such an approach on several types of circadian data.
First, we show using single-cell fibroblast data that stochastic noise is directly related to population-level damping.
Next, we show that a small-molecule modulator is able to change damping rate in a dose-dependent fashion, and verify using a mathematical model that changes to single-cell stochasticity is a likely mechanism.
Finally, we calculate the genome-wide effects of siRNA knockdown on single-cell noise, and demonstrate that population-level damping rate is independent of other circadian parameters, such as period or amplitude.
Using this additional information, we show that circadian rhythms have likely evolved to an optimal point of high amplitude and low stochastic noise.
Our results should prove especially important in the future search for small molecule circadian therapeutics, as it allows the effect of candidate drugs on single-cell noise to be quantified in a high-throughput manner.

@@

\subsubsection*{Sorting cells by noise level}
As in the original study, various parameters describing the average noise level of each cell were collected.
Traces were denoised and detrended by keeping only the $(8 \text{hr}, 258 \text{hr})$ wavelet components.
From these smoothed trajectories, a Hilbert transform was used estimate points at which the phase crossed $0$ to find period and amplitude coefficients of variation.
An additional noise parameter, the standard deviation in the $(1 \text{hr}, 8 \text{hr})$ wavelet components divided by the overall rhythm amplitude, was used to quantify the high-frequency noise of the system.
From these three noise variables, a combined noise metric was constructed by projecting the variables along their first principle component (using scikit-learn, \url{http://scikit-learn.org/}).
Cells were ranked according to this combined noise metric, and a high-noise group and low-noise group were constructed by taking the 39 highest-noise and lowest-noise cells, respectively.
The raw bioluminescence profiles were not initially synchronized, so in order to simulate the gradual desynchronization of a group of oscillators the traces were offset to have the same starting phase. 
This was accomplished by starting each trace at the first phase zero-crossing, which were found using a Hilbert transform.

\subsubsection*{Bootstrap estimations of the damping rate difference}
Averaged traces for low and high-noise group displayed a damped sinusoidal rhythm.
The first 4 days of rhythms ($s = 0.5,\ N = 192$) were fit using a damped sinusoid.
To ensure the difference in damping rate between groups was statistically significant, a bootstrap analysis was performed.
In each of $10,000$ bootstrap trials, the cells were randomly assigned evenly to either the low-noise or the high-noise group (one cell was omitted randomly in each trial to ensure even group sizes).
The absolute difference in damping rate between the two populations was recorded to yield a two-tailed test.
The observed test statistic, $|d_h - d_l| = 6.65\times10^{-3}$, was found to be significant at the $\alpha = 0.05$ confidence threshold ($p = 0.0264$).

\subsection*{Quantifying dose-dependent effects of small molecule modulators}
Bioluminescence traces ($s = 1.67,\ N=71$) from the application of increasing concentrations of small molecule were fit using the method described in a previous section.
Because the small molecules were toxic to the cells at very high concentrations, experiments were removed from further analysis when the $R^2$ of the sinusoidal regression fell below $0.80$ (Figure~\ref{fig:small_molecule_r2}).

\subsection*{{In silico} prediction of small molecule experiments}
A previously published mathematical model of circadian rhythms was used to predict the effects on population damping rate from the dose-dependent small molecule experiments \cite{Hirota2012}.
The parameters used to capture the effects of each small molecule were the same as described previously \cite{St.John2014}.
The model was converted to a stochastic biochemical system and subsequently simulated using StochKit2 \cite{Sanft2011a} (via GillesPy, \url{https://github.com/JohnAbel/gillespy}).
Population-level rhythms were found by taking the average of $1,000$ noninteracting oscillators, starting at identical initial conditions.
The only parameter left unspecified by the deterministic model was the cell volume $\Omega$, which controlled the amount of noise in the system.
For a given $\Omega$, the population-level damping rate was found by finding the phase diffusivity parameter which gave the closest fit to the simulated population-level rhythms, as described in \cite{St.John2014a}.
An $R^2$ value was calculated for each fit, taking into account all eight state variables.

\subsubsection*{Fitting the volume parameter}
We calculated an average experimental damping rate of $d = 0.0151$ from the $0 \mu M$ bioluminescence trajectories for both KL001 and longdaysin.
{\itshape In silico} damping rates were calculated for logarithmically space values of $\Omega \in (100, 500)$.
Ten independent groups of $1,000$ oscillators were simulated for each $\Omega$, from which the means and standard errors were found.
Simulations in which $R^2 < 0.90$ were removed from further analysis.
A weighted least-squares regression (using statsmodels, \url{http://statsmodels.sourceforge.net/}) was performed for $\log d$ vs.\ $\log \Omega$, using the $\log \text{SEM}$ of each measurement as a regression weight (Figure~\ref{fig:vol_calibration}).
The best fit was found to be $\Omega = 226.3 \pm 9.0$.

@@

\subsection*{Fitting the genome-wide siRNA screen}
We analyzed the data and annotations for the $111,743$ wells ($s=2,\ N=72$) in the Zhang {\itshape et al.,} 2009 screen \cite{Zhang2009}.
Fits for which the $R^2 < 0.80$ were discarded.
We analyze the natural logarithm of the amplitude, as these are more normally distributed and on a similar scale to the damping rate.
Plate to plate variation, as shown in Figure~\ref{fig:plate_variation}, was more severe than variation on a well-to-well basis, Figure~\ref{fig:well_variation}.
Parameters were therefore normalized on a plate-by-plate basis using a robust z-score \cite{Birmingham2009}:
\[
  z_{R,i} = \frac{p_i - M(p_i)}{M\left(\left|p_i - M(p_i)\right|\right)}, \qquad i \in \{0, \dots, {\cal P} - 1 \}
\]
where $M(\cdot)$ denotes the median of a vector, and $p_i$ contains all the points in one plate, and $\cal P$ is the number of plates.
We removed outlier points prior to calculating the moments of the distributions, Pearson correlation coefficients, and performing the multivariable linear regression.
Outliers were defined as points which contained a z-metric (in either period, phase, amplitude, or damping rate) with an absolute value greater than eight.
We chose the ``control'' wells to be those which contained no siRNA, as these proved to be more numerous than those containing reference siRNA perturbations and were clustered similarly to the highest-density regions of the perturbed fits.

\subsubsection*{Detecting outlier perturbations}
We found the average response to each gene perturbation by grouping the perturbed dataset by target gene ID (using pandas, \url{http://pandas.pydata.org/}).
A Hotelling's $T^2$ test was used to determine whether the means of each gene knockdown was significantly different from the control population.
A robust covariance estimator was used to find the location and covariance of the control distribution (using scikit-learn).
Because the control distribution ($n=11,253$) is much larger than that for any particular gene ID ($n \approx 4$), we assume the pooled covariance can be approximated by the covariance of the control sample alone.

\subsection*{Software}
All computations were performed using Python.
Codes used to produced the figures in this manuscript can be found at \url{https://github.com/pstjohn/decay-methods}.

@@

We first sought to determine whether single-cell stochastic noise affects the population-level damping rate in cultured bioluminescence cells.
To do this, we calculated noise characteristics from experimental data on individual PER2::LUC fibroblast cells \cite{Leise2012}.
We sorted the cells into two groups, a low-noise group and a high-noise group, based on the relative high-frequency noise, period variability, and amplitude variability present in each trace.
Example rhythms from cells in both groups are shown in Figure~\ref{fig:fibroblast_noise}A.
Because the cells were not synchronized at the start of the recording, we replicate this effect {\itshape in silico} by shifting each series in time to align their start phases.
We next calculated expected population-level bioluminescence traces by averaging the cellular PER2::LUC signal in each group.
Both populations displayed averaged rhythms which resembled a damped sinusoid, similar to those seen in bioluminescence recordings of entire cell cultures.
Fitting the averaged expression of each group with a damped sinusoid revealed that the low-noise group also had a lower damping rate (Figure~\ref{fig:fibroblast_noise}B).
The significance of this difference was confirmed via a bootstrap analysis (Figure~\ref{fig:fibroblast_noise}C), where cells were randomly assigned in each bootstrap trial to either the low-noise or the high-noise group.

@@

We next demonstrate that perturbations to the transcriptional oscillator are capable of altering single-cell level stochastic noise.
The actions of small-molecule circadian modulators KL001 and longdaysin are well-characterized, and are known to affect circadian period and amplitude in a dose-dependent fashion \cite{St.John2014}.
By fitting experimental data on the population-level responses to increasing dosages of each molecule with a damped sinusoid, we show that KL001, but not longdaysin, increases damping rate in a dose-dependent fashion (Figure~\ref{fig:dose_dependence}).
This change in damping rate is consistent across both reporter systems ({\itshape Bmal1-dLuc} and {\itshape Per2-dLuc}), indicating it is a fundamental property of the overall gene regulatory network.

@@

\begin{figure}[tbp]
  \begin{center}
    \includegraphics[width=0.8\textwidth]{figures/pdfs/fitted_parameters.pdf}
  \end{center}
  \caption{{\bfseries Distributions in fitted parameters for the genome-wide siRNA screen.} Distributions in robust z-scores closely resemble normal distributions. For all parameters, the region of highest density is consistent between the control and perturbed populations, indicating many perturbations do not appreciably change clock dynamics. } 
\label{fig:fit_distributions}
\end{figure}

@@

Since an siRNA's effect on period length does not effectively predict its effect on amplitude or phase (and vice-versa), we hypothesized that stochastic noise would similarly be independently affected by siRNA perturbations.
Low Pearson correlation coefficients between normalized parameter distributions defend this hypothesis, with the highest correlations among variables seen between amplitude and damping rate ($\rho = 0.285$, Table~\ref{tab:corr}).
Additionally, a multivariate linear regression of damping rate as a function of period, amplitude, phase, and perturbation type (control or perturbed, a categorical variable) produced an $R^2$ value of only $0.169$ (Table~\ref{tab:ols_reg}).

\begin{table}
  \begin{center}
    \begin{tabular}{rrrrr}
      \toprule
      {}       & $d$    & $\ln A$ & $T$    & $\theta$ \\\midrule
      $d$      & $1.000 $ & $0.285 $  & $-0.142$ & $-0.269$\\\
      $\ln A$  & $0.285 $ & $1.000 $  & $-0.022$ & $-0.112$\\
      $T$      & $-0.142$ & $-0.022$  & $1.000 $ & $-0.113$\\
      $\theta$ & $-0.269$ & $-0.112$  & $-0.113$ & $1.000 $\\
      \bottomrule
    \end{tabular}
  \end{center}
  \caption{{\bfseries Correlation among normalized parameters of the high-throughput siRNA screen.} Pearson correlation coefficients are relatively low between fitted parameters, indicating that changes to damping rate (and thereby stochastic noise) are not explained by changes to period, amplitude, or phase.}
  \label{tab:corr}
\end{table}

@@

The ability of a population of oscillators to maintain robust oscillations is a function both of its initial amplitude as well as its damping rate.
A robust oscillator is therefore one with high amplitude and low stochastic noise.
Scatter plots of the control and perturbed distributions are shown in Figure~\ref{fig:outlier_dist}A-B, which indicate that few outlier points reside in the lower damping rate, higher amplitude quadrant.
In order to find perturbations which confidently change robustness, we grouped the siRNA perturbations by target gene and performed a two-sample Hotelling's $T^2$ test against the control population.
The resulting significant gene perturbations ($75\%$ of all genes) are shown in Figure~\ref{fig:outlier_dist}C.
Quantifying the distribution in outliers by quadrant, it is clear that only a small minority of perturbations ($3.3\%$) simultaneously increase amplitude and decrease stochastic noise.
It therefore appears that the unperturbed clock optimizes some trade-off between oscillator amplitude and stochastic noise, such that it is not feasible to make oscillations more robust by knocking down one gene at a time.

@@

@@

In this study, we have described a method by which changes to stochastic noise can be estimated from population-level circadian bioluminescence recordings.
Our method assumes independent and sustained oscillators at the single-cell level.
While these assumptions have not always been upheld in the circadian literature \cite{Izumo2003}, more recent studies have demonstrated the validity of these assumptions in cultured fibroblast cells.
Earlier mathematical studies have analysed on the damping rate of oscillators without assuming independent or homogeneous populations \cite{Rougemont2007}.
However, such approaches introduce many additional fitted parameters, limiting their applicability to high-throughput screens.

>>>>

\pagebreak
\begin{abstract}
Stochastic noise at the cellular level has been shown to play a fundamental role in circadian oscillations, influencing how groups of cells entrain to external cues and likely serving as the mechanism by which cell-autonomous rhythms are generated.
Despite this importance, few studies have investigated how clock perturbations affect stochastic noise - even as increasing numbers of high-throughput screens categorize how gene knockdowns or small molecules can change clock period and amplitude.
This absence is likely due to the difficulty associated with measuring cell-autonomous stochastic noise directly, which currently requires the careful collection and processing of single-cell data.
% since cultured circadian reporters lack significant intercellular coupling,
In this study, we show that the damping rate of population-level bioluminescence recordings can serve as an accurate measure of cell-autonomous stochastic noise, and one which can be applied to future and existing high-throughput circadian screens.
Using cell-autonomous fibroblast data, we first show directly that higher noise at the single-cell results in faster damping at the population level.
Next, we show that the damping rate of cultured cells can be changed in a dose-dependent fashion by small molecule modulators, and confirm that such a change can be explained by single-cell noise using a mathematical model.
We further demonstrate the insights that can be gained by applying our method to a genome-wide siRNA screen, revealing that stochastic noise is altered independently from period, amplitude, and phase.
Finally, we hypothesize that the unperturbed clock is highly optimized for robust rhythms, as very few gene perturbations are capable of simultaneously increasing amplitude and lowering stochastic noise.
Ultimately, this study demonstrates the importance of considering the effect of circadian perturbations on cell-autonomous noise, particularly with regard to the development of small-molecule circadian therapeutics.
\end{abstract}

@@

Circadian rhythms are daily changes in gene expression and physiology that persist even in the absence of external environmental cues \cite{Herzog2007}.
In mammals, such rhythms are organized in a hierarchical fashion: at the tissue-level, the brain's suprachiasmatic nucleus (SCN) serves as the master pacemaker and keeps circadian oscillations in peripheral tissues in phase with the light-dark cycle.
Cell-to-cell coupling in the SCN keeps individual cells in tight synchrony \cite{Herzog2004}. 
Within each tissue, cellular-level rhythms in gene transcription are generated by a large network of interacting regulatory elements, in which time-delayed negative feedback gives rise to sustained oscillations \cite{Ueda2005}.
The robust oscillation of circadian factors has been linked to metabolic health \cite{Bass2012}, since rhythms compromised by gene knockout \cite{Marcheva2010} or irregular feeding schedules \cite{Hatori2012} result in an increased risk of metabolic disease.
Additionally, as the amplitude of circadian transcription can be affected by lifestyle variables such as diet, age, or work schedules, there has been recent interest in developing pharmacological strategies for increasing the amplitude of circadian cycles in metabolic tissues \cite{Chen2013}.

A detailed understanding of the underlying transcriptional mechanisms is essential for the development of circadian therapeutics to be successful. 
The functional roles of different genes in circadian regulation have traditionally been studied using behavioral-level data and genetic knockout experiments \cite{Vitaterna1994}.
Bioluminescence-based cellular circadian reporters offer a more direct view of the gene regulatory network \cite{Balsalobre1998} and are amenable to high-throughput screens, allowing genome-wide exploration into factors which affect circadian rhythmicity \cite{Zhang2009}.
Additionally, cultured circadian reporter cells allow the change in transcriptional amplitude following a perturbation to be quantified.
This additional parameter has proven useful in differentiating between perturbations with the same effect on period \cite{St.John2014} and has aided the search for small-molecule therapeutics to boost clock amplitude \cite{Chen2013}.

@@

Despite the averaging which occurs at the population-level, cell-autonomous stochastic noise plays an important role in determining the function of the circadian oscillator.
A recent study showed that stochasticity is critical to the population-level response to a neuropeptide and forms the basis for how the SCN entrains to light-mediated cues \cite{An2013}.
Additional studies have argued that the basis of single-cell rhythmicity depends on stochastic noise, as models of deterministically damped oscillators, when simulated stochastically, better capture the noise characteristics seen in single-cell fibroblast data \cite{Westermark2009}.
Despite the importance of single-cell stochasticity in circadian rhythms, measuring stochastic noise currently requires careful preparation, recording, and image processing of individual cells \cite{Leise2012}. 
As a result, while circadian perturbations have been postulated to affect single-cell stochasticity \cite{Rougemont2007}, no study has experimentally quantified changes to single-cell noise as a result of a small molecule or genetic perturbation.

In this study, we demonstrate that changes to single-cell stochasticity can be reliably inferred from the changes to the damping rate of population-level bioluminescence recordings of cultured circadian reporters.
Our method assumes that oscillations in individual cells are both
\begin{itemize}
  \item independent (no significant cell-to-cell coupling) \cite{Noguchi2013, Guenthner2014},
  \item and sustained (do not damp on a single-cell basis) \cite{Welsh2004, Nagoshi2004},
\end{itemize}
which have been shown to hold experimentally for cultured fibroblast cells.
We demonstrate the validity and usefulness of such an approach on several types of circadian data.
First, we show using single-cell fibroblast data that stochastic noise is directly related to population-level damping.
Next, we show that a small-molecule modulator is able to change damping rate in a dose-dependent fashion, and verify using a mathematical model that changes to single-cell stochasticity is a likely mechanism.
Finally, we calculate the genome-wide effects of siRNA knockdown on single-cell noise, and demonstrate that population-level damping rate is independent of other circadian parameters, such as period or amplitude.
Using this additional information, we show that circadian rhythms have likely evolved to an optimal point of high amplitude and low stochastic noise.
Our results should prove especially important in the future search for small molecule circadian therapeutics, as it allows the effect of candidate drugs on single-cell noise to be quantified in a high-throughput manner.

@@

\subsubsection*{Sorting cells by noise level}
As in the original study, various parameters describing the average noise level of each cell were collected.
Traces were denoised and detrended by keeping only the $(8 \text{hr}, 258 \text{hr})$ wavelet components.
From these smoothed trajectories, a Hilbert transform was used estimate points at which the phase crossed $0$ to find period and amplitude coefficients of variation.
An additional noise parameter, the standard deviation in the $(1 \text{hr}, 8 \text{hr})$ wavelet components divided by the overall rhythm amplitude, was used to quantify the high-frequency noise of the system.
From these three noise variables, a combined noise metric was constructed by projecting the variables along their first principle component (using scikit-learn, \url{http://scikit-learn.org/}).
Cells were ranked according to this combined noise metric, and a high-noise group and low-noise group were constructed by taking the 39 highest-noise and lowest-noise cells, respectively.
The raw bioluminescence profiles were not initially synchronized, so the traces were offset to have the same starting phase in order to simulate the gradual desynchronization of a group of oscillators. 
This was accomplished by starting each trace at the first phase zero-crossing, found using a Hilbert transform.

\subsubsection*{Bootstrap estimations of the damping rate difference}
Averaged traces for low and high-noise group displayed a damped sinusoidal rhythm.
The first 4 days of rhythms ($T_s = 0.5,\ N = 192$) were fit using a damped sinusoid.
To ensure the difference in damping rate between groups was statistically significant, a bootstrap analysis was performed.
In each of $10,000$ bootstrap trials, cells were randomly assigned evenly to either the low-noise or the high-noise group (one cell was randomly omitted in each trial to ensure even group sizes).
The absolute difference in damping rate between the two populations was recorded to yield a two-tailed test.
The observed test statistic, $|d_h - d_l| = 6.65\times10^{-3}$, was found to be significant at the $\alpha = 0.05$ confidence threshold ($p = 0.0264$).

\subsection*{Quantifying dose-dependent effects of small molecule modulators}
Bioluminescence traces ($T_s = 1.67,\ N=71$) with increasing small molecule concentrations were fit with a damped sinusoid using the method described in a previous section.
Because the small molecules were toxic at very high concentrations, experiments were removed from further analysis where $R^2 < 0.80$ (Figure~\ref{fig:small_molecule_r2}).

\subsection*{{In silico} prediction of small molecule experiments}
A previously published mathematical model of circadian rhythms \cite{Hirota2012} was used to predict the effects on population damping rate from the dose-dependent small molecule experiments.
The parameters used to capture the effects of each small molecule were the same as described previously \cite{St.John2014}.
The model was converted to a stochastic biochemical system and subsequently simulated using StochKit2 \cite{Sanft2011a} (via GillesPy, \url{https://github.com/JohnAbel/gillespy}).
Population-level rhythms were found by taking the average of $1,000$ noninteracting oscillators, starting at identical initial conditions.
The only parameter left unspecified by the deterministic model was the cell volume, $\Omega$, which controlled the amount of noise in the system.
For each $\Omega$, a damped sinusoid was fit to the population-averaged state trajectory. 
An $R^2$ value was calculated for each fit, taking into account all eight state variables.

\subsubsection*{Fitting the volume parameter}
We calculated an average experimental damping rate of $d = 0.0151$ from the $0 \mu M$ bioluminescence trajectories for both KL001 and longdaysin.
{\itshape In silico} damping rates were calculated for logarithmically spaced values of $\Omega \in (100, 500)$.
Ten independent groups of $1,000$ oscillators were simulated for each $\Omega$, from which the means and standard errors were found.
Simulations in which $R^2 < 0.90$ were removed from further analysis.
A weighted least-squares regression (using statsmodels, \url{http://statsmodels.sourceforge.net/}) was performed for $\log d$ vs.\ $\log \Omega$, using the $\log \text{SEM}$ of each measurement as a regression weight (Figure~\ref{fig:vol_calibration}).
The best fit was found to be $\Omega = 226.3 \pm 9.0$, and was used for subsequent model simulations.

@@

\subsection*{Fitting the genome-wide siRNA screen}
We analyzed the data and annotations for the $111,743$ wells ($T_s = 2,\ N = 72$) in the Zhang {\itshape et al.,} 2009 screen \cite{Zhang2009}.
Fits for which the $R^2 < 0.80$ were discarded.
The natural logarithm of the amplitude was used, since it more closely resembled a normal distribution and was on a similar scale to the damping rate.
Plate to plate variation, as shown in Figure~\ref{fig:plate_variation}, was more severe than variation on a well-to-well basis, Figure~\ref{fig:well_variation}.
Parameters were therefore normalized on a plate-by-plate basis using a robust z-score \cite{Birmingham2009}:
\[
  z_{R,i} = \frac{p_i - M(p_i)}{M\left(\left|p_i - M(p_i)\right|\right)}, \qquad i \in \{0, \dots, {\cal P} - 1 \}
\]
where $M(\cdot)$ denotes the median of a vector, and $p_i$ contains all the points in one plate, and $\cal P$ is the number of plates.
We removed outlier points prior to calculating the moments of the distributions, Pearson correlation coefficients, and performing the multivariable linear regression.
Outliers were defined as points which contained a z-metric (in either period, phase, amplitude, or damping rate) with an absolute value greater than eight.
We chose the ``control'' wells to be those which contained no siRNA, as these proved to be more numerous than those containing reference siRNA perturbations and were clustered similarly to the highest-density regions of the perturbed fits.

\subsubsection*{Detecting outlier perturbations}
We found the average response to each gene perturbation by grouping the perturbed dataset by target gene ID (using pandas, \url{http://pandas.pydata.org/}).
A Hotelling's $T^2$ test was used to determine whether the means of each gene knockdown was significantly different from the control population.
A robust covariance estimator was used to find the location and covariance of the control distribution (using scikit-learn).
Because the control distribution ($n=11,253$) is much larger than that for any particular gene ID ($n \approx 4$), the effect of the perturbed sample on the pooled covariance was neglected.

\subsection*{Software}
All computations were performed using Python.
Code used to perform the analysis and produce the figures in this manuscript can be found online at \url{https://github.com/pstjohn/decay-methods}.

@@

We first sought to determine whether changes to single-cell stochastic noise affect population-level damping rate.
To do this, we calculated noise characteristics from experimental data on individual PER2::LUC fibroblast cells \cite{Leise2012}.
Cells were sorted into two groups, a low-noise group and a high-noise group, based on the relative high-frequency noise, period variability, and amplitude variability present in each trace.
Example rhythms from cells in both groups are shown in Figure~\ref{fig:fibroblast_noise}A.
Because the cells were not synchronized at the start of the recording, this effect is replicated {\itshape in silico} by shifting each series in time to align their start phases.
Population-level bioluminescence traces were then found by averaging the cellular PER2::LUC signal in each group.
Both populations displayed averaged rhythms which resembled a damped sinusoid, similar to those seen in bioluminescence recordings of entire cell cultures.
Fitting the averaged expression of each group with a damped sinusoid revealed that the low-noise group also had a lower damping rate (Figure~\ref{fig:fibroblast_noise}B).
The significance of this difference was confirmed via a bootstrap analysis (Figure~\ref{fig:fibroblast_noise}C), where cells were randomly assigned in each bootstrap trial to either the low-noise or the high-noise group.

@@

We next demonstrate that perturbations to the transcriptional oscillator are capable of altering single-cell level stochastic noise.
The actions of small-molecule circadian modulators KL001 and longdaysin are well-characterized, and are known to affect circadian period and amplitude in a dose-dependent fashion \cite{St.John2014}.
By fitting experimental data on the population-level responses to increasing dosages of each molecule with a damped sinusoid, we show that KL001, but not longdaysin, increases damping rate in a dose-dependent fashion (Figure~\ref{fig:dose_dependence}).
This change in damping rate is consistent across both reporter systems ({\itshape Bmal1-dLuc} and {\itshape Per2-dLuc} U2OS cells), indicating it is a fundamental property of the overall gene regulatory network.

@@

\begin{figure}[tbp]
  \begin{center}
    \includegraphics[width=0.8\textwidth]{figures/pdfs/fitted_parameters.pdf}
  \end{center}
  \caption{{\bfseries Distributions in fitted parameters for the genome-wide siRNA screen.} Distributions in robust z-scores closely resemble normal distributions. For all parameters, the region of highest density is consistent between the control and perturbed populations, indicating many perturbations do not appreciably change clock dynamics.} 
\label{fig:fit_distributions}
\end{figure}

@@

Since an siRNA's effect on period length does not effectively predict its effect on amplitude or phase (and vice-versa), we hypothesized that stochastic noise would similarly be independently affected by siRNA perturbations.
Low Pearson correlation coefficients between normalized parameter distributions defend this hypothesis, with the highest correlations among variables seen between amplitude and damping rate ($\rho = 0.285$, Table~\ref{tab:corr}).
Additionally, a multivariate linear regression of damping rate as a function of period, amplitude, phase, and perturbation type (control or perturbed, a categorical variable) produced an $R^2$ value of only $0.169$ (Table~\ref{tab:ols_reg}).
These results reinforce the notion that the population-level damping rate describes an independent feature of the underlying oscillator.

\begin{table}
  \begin{center}
    \begin{tabular}{rrrrr}
      \toprule
      {}       & $d$    & $\ln A$ & $T$    & $\theta$ \\\midrule
      $d$      & $1.000 $ & $0.285 $  & $-0.142$ & $-0.269$\\
      $\ln A$  & $0.285 $ & $1.000 $  & $-0.022$ & $-0.112$\\
      $T$      & $-0.142$ & $-0.022$  & $1.000 $ & $-0.113$\\
      $\theta$ & $-0.269$ & $-0.112$  & $-0.113$ & $1.000 $\\
      \bottomrule
    \end{tabular}
  \end{center}
  \caption{{\bfseries Correlation among normalized parameters of the high-throughput siRNA screen.} Pearson correlation coefficients are relatively low between fitted parameters, indicating that changes to damping rate (and thereby stochastic noise) are not explained by changes to period, amplitude, or phase.}
  \label{tab:corr}
\end{table}

@@

The ability of a population of oscillators to maintain robust oscillations is a function both of its initial amplitude as well as its damping rate.
A robust oscillator is therefore one with high amplitude and low stochastic noise.
Scatter plots of the control and perturbed distributions are shown in Figure~\ref{fig:outlier_dist}A-B, which indicate that few outlier points reside in the lower damping rate, higher amplitude quadrant.
In order to find perturbations which confidently change robustness, we grouped the siRNA perturbations by target gene and performed a two-sample Hotelling's $T^2$ test against the control population.
The resulting significant gene perturbations ($75\%$ of all genes) are shown in Figure~\ref{fig:outlier_dist}C.
Quantifying the distribution in outliers by quadrant, it is clear that only a small minority of perturbations ($3.3\%$) simultaneously increase amplitude and decrease stochastic noise.
It therefore appears that the unperturbed clock optimizes some trade-off between high amplitude and low stochastic noise, such that it is unlikely the knockdown of a single gene will make oscillations more robust.

@@

\subsection*{Effect of cell heterogeneity}

Experimental work has shown that cell-autonomous fibroblast cells have a distribution of free-running periods \cite{Leise2012}.
Indeed, prior to the availability of single-cell data, studies explored the possibility that differences in mean periods served as the mechanism behind population-level damping \cite{Izumo2003}.
While it is true that the dephasing of a group of oscillators can be caused by both variance in the mean period as well as cycle-to-cycle variability, in the preceding analyses we neglect the effect of period heterogeneity.
To motivate this assumption, we show that there is greater variance in period on a cycle-to-cycle basis than on a cell-to-cell basis (Figure~\ref{fig:cell2cell-variability}).
Additionally, it is important to note that the dephasing kinetics of these two alternative processes are distinguishable from population-level data.
In dephasing due solely to stochastic noise, the probability density of finding a cell with any given phase follows a standard convection-diffusion regime, resulting in exponentially damped rhythms \cite{St.John2014a}.
In dephasing due solely to period heterogeneity, the variance of the phase probability density broadens in time proportionally to $t^2$, rather than $t$ in standard diffusion.
This type of anomalous diffusion results in a different damping profile, as highlighted in Figure~\ref{fig:dephasing-rates}.
Since the data used in this study is well-described by an exponentially damped sinusoid, and since a mathematical model assuming homogeneous mean periods accurately predicts the experimental damping rate changes, this assumption is likely justified.
Additionally, since this study is mainly concerned with changes in the damping rate (rather than the absolute damping rate), it is unlikely that period heterogeneity plays a major role.

@@

% Our method assumes independent and sustained oscillators at the single-cell level.
% While these assumptions have not always been upheld in the circadian literature \cite{Izumo2003}, more recent studies have demonstrated the validity of these assumptions in cultured fibroblast cells.
% Earlier mathematical studies have analysed on the damping rate of oscillators without assuming independent or homogeneous populations \cite{Rougemont2007}.
% However, such approaches introduce many additional fitted parameters, limiting their applicability to high-throughput screens.
